Non-parametric approaches to optimal policy are more robust

### 2013-03-21 inst/doc/nonparametric-bayes.Rmd
long way to go

@@

Structural uncertainty.

@@

```{r echo=FALSE, results="asis"} #bibliography("html") ```

>>>>

Process-based models built on modern machine learning approaches 
offer a powerful and underappreciated approach for ecological
forecasting and management.  We present a novel example of how
one such example - non-parametric Gaussian processes, can be adapted
to the context of management and decision-making.

While effective machine learning algorithms have speckled the literature, 
most theoretical ecologists in particular have been wary of these 'black-box'
approaches in favor of simple, mechanistic models.  While a blind application
of machine learning approaches would indeed be dangerous grounds to stake 
either ecological understanding or management practice, such concerns are not
grounds to reject such methods out of hand.  Rather, further attention is 
warranted to identify the circumstances in which machine-learning approaches
will most likely outperform existing more mechanistic approaches, or help
crack difficult management questions where ecological or dynamical complexity
has caused traditional approaches to founder.

Here we illustrate how a machine learning model can be implemented in place
of simple mechanistic model in a decision theory (optimal control) problem,
and highlight the features of this system that make the nonparametric approach
particularly effective.

<!-- Um, Intro take 2, here we go... -->

Mechanistic models have long been the gold standard of theoretical
modeling in ecology (e.g. see @Geritz2012 or @Cuddington2013).  Only by
understanding the processes involved can we make reliable long term
predictions and build an knowledge of cause and effect that guides the
hypotheses we make, the data we collect, and the management decisions
we make.  Process-based models, whether expressed in the language of
mathematics or English, identify the connection between mosquitoes and
the spread of malaria, or greenhouses gases and climate change, guiding
our approach to understand and manage these threats. Despite this central
importance, we argue here that ecologists would do well to give greater
attention to the role non-mechanistic models can play in ecological
management and decision making.  The value of these approaches is
greatest in a context where decisions are made over short time horizons,
and updated as new data becomes available.

## The alternative to process-based models: pattern-based modeling

Before we proceed further it would be useful to define our terms and lay
some greater context for the discussion. Throughout, we will distinguish
"mechanistic" or "process-based" models from models that are merely
"correlative" or "pattern-based". Historically pattern-based modeling
meant regression (usually linear regression) and resided in methodology
rooted in statistics departments and was the primary focus of empirical
ecologists, while mechanistic modeling meant dynamical systems (ODEs,
PDEs, SDEs and their discrete kin), residing in methodology from
the mathematics department and was the primary focus of theoretical
ecologists.

We believe these historical divides continue to color much of the
literature today, with theorists more skeptical than empiricists of
statistical methodology and vice versa.  Meanwhile, the ground underneath
has shifted.  Originally pattern-matching approaches could be critiqued
on the grounds of their simplicity (not everything is linear) while
mechanistic approaches could be critiqued on their tenuous connections
to data -- model parameter values such as death rates, birth rates,
etc. would be estimated in advance and then stuck into the model, rather
than estimated in the context of the model itself.  Today both of these
critiques are outdated. Computational power and hierarchical statistical
methods (particularly approximate likelihood or simulation techniques such
as particle filters and approximate Bayesian computing) have brought the
inference richer dynamical systems into their fold, while pattern-based
modeling has spawned an entirely new approach under the banner of machine
learning that has divided the statistics community (See @Bremman2001,
"The Two Cultures").   Machine learning models can represent almost
arbitrarily complex patterns and incorporate learning and decision making
strategies that have earned them the name 'artificial intelligence.'
Yet at the same time they can be far less transparent their linear
regression predecessors, making both astounding successes and startling
biases and failures like their biological namesake. More ecologists,
particularly theorists and modelers, would do well to learn both how to
take advantage of these strengths and recognize their weaknesses.

Mechanistic modeling emphasizes the importance of capturing the correct
gross properties of a system over tracking minute fluctuations.
For instance, in selecting and parameterizing model of a population
of conservation concern, we may be most interested in getting the
long-term behavior correct -- such as identifying if the dynamics support
persistence of the population -- rather than worrying how well they
reflect the year-to-year fluctuations.  We would certainly have good
reason to prefer such a model over alternatives which are irreconcilable
to the most basic biological processes, such as unbounded growth, or
growth curves that do not pass through the origin in a closed system. So
it may come as a surprise to realize that such obviously wrong models
can perform as well or even better than reasonable mechanistic models
in guiding ecological management and decision making.

<!--- Um, Intro take 3? here we go again... -->

@@

@@

```{r echo=FALSE, results="asis"}
#bibliography()
```

### 2013-03-21 inst/doc/nonparametric-bayes.tex
long way to go

The problem of structural uncertainty in managing ecological systems.

(Big literature here of course, primary approaches have been using
parametric uncertainty and model uncertainty (e.g.~belief distribution
over models I believe. Worth distinguishing our problem from the more
trivially sized problems in which the action space is discrete and
consists of 2 or three possible choices, e.g.~much of the recent
Possingham et al work.)

\begin{itemize}
\item
  Most management recommendations from the ecological literature /
  management policies based on (motivated by) parametric models.
  \emph{Preference to pitch towards policy or theoretical literature?}
\item
  Background discussion on the importance/success of parametric
  modeling; (e.g. Levins (1966), etc., up through Geritz and Kisdi
  (2011))
\item
  Background on the concerns of structural uncertainty -- we don't have
  the right models. (Also: measurement uncertainty, parameter
  uncertainty, unobserved states, boundary conditions, etc.)
\item
  Hierarchical Bayesian approach has provided a natural way to address
  these from a statistical standpoint. Successes and challenges from the
  parametric route: e.g. Cressie et al. (2009).
\item
  Among pitfalls of these approaches: particularly difficult to apply in
  management context. Optimization-based (Decision-theoretic) approaches
  to ecological management need to be precisely parameterized about
  everything, even the uncertainty itself (Polasky et al. 2011)
  (discussion of Brozović and Schlenker (2011) SDP example in context of
  threshold system would perhaps be useful as well).
\item
  Management goals / decision-theoretic approaches need accurate
  prediction over relevant (short?) timescales more than accurate (but
  incomplete or noisy estimated) mechanisms.
\item
  Nonparametric (machine-learning?) approaches may offer the benefit of
  the hierarchical Bayesian approach without the practical and
  computational limitations of their parametric kin. Non-parametric
  models are flexible enough to take maximum advantage of the data
  available, while being appropriately ambiguous about the dynamics of a
  system in regions of parameter space that have been poorly or never
  sampled.
\item
  Nonparametric approaches are beginning to appear more frequently in
  ecological and conservation literature (Species distribution
  models/maxent), including the Gaussian process based approach used
  here (Munch et al. 2005). However, such approaches have yet to be
  applied to the decision-theoretic framework that could guide
  management decisions, where we expect them to excel for several
  reasons (1) the ability to make accurate forecasts by more closely
  approximating the underlying process where data is available (2)
  remaining appropriately ambiguous where data is not available (3)
  remaining computationally simple enough to avoid some pitfalls common
  to hierarchical parametric approaches.
\end{itemize}

@@

The use of Gaussian process regression (known as Kreging in the
geospatial literature) to formulate a predictive model is relatively new
in the context of modeling dynamical systems {[}@Kocijan2005{]} and
introduced in the ecological modeling and fisheries management by Munch
et al. (2005).

\emph{Is the following discription useful or too elementary?}

The essense of the Gaussian process approach can be captured in the
following thought experiment: An exhaustive parametric approach to the
challenge of structural uncertainty might proceed by writing down all
possible functional forms for the underlying dynamical system with all
possible parameter values for each form, and then consider searching
over this huge space to select the most likely model and parameters; or
using a Bayesian approach, assign priors to each of these possible
models and infer the posterior distribution of possible models. The
Gaussian process approach can be thought of as a computationally
efficient approximation to this approach. Gaussian processes represent a
large class of models that can be though of as capturing or reasonably
approximating the set of models in this collection. By modeling at the
level of the process, rather than the level of parametric equation, we
can more concisely capture the possible behavior of these curves. In
place of a parametric model of the dynamical system, the Gaussian
Process approach postulates a prior distribution of (n-dimensional)
curves that can be though of as approximations to a large range of
possible

\emph{Do we need more specifics on Gaussian process as an approximation
to parametric models? Discussion of Gaussian process vs other machine
learning / forecasting approaches that have less clear statistical
foundations? If so, does this all belong in the discussion?}

The Gaussian process is defined by a covariance kernel \ldots{}
\emph{discussion of our (gaussian) kernel, blah blah how specific here
vs appendix?}. By requiring our kernel to follow a generic functional
form, we can compactly describe the Gaussian process using only a
handful of parameters (Table 1)

A more thorough introduction to the formulation and use of Gaussian
processes can be found in @Rasmussen2006.

\begin{itemize}
\item
  Gaussian processes in ecological literature (Munch et al. 2005)
\item
  Our methodology (e.g.~following Munch et al. (2005)).
\end{itemize}

\subsubsection{Discussion of the dynamic programming solution}

@@

\begin{itemize}
\item
  Non-parametric methods have received far too little attention in
  ecological modeling efforts that are aimed at improved conservation
  planning and decision making support.
\item
  Importance of non-parametric approaches in conservation planning /
  resource management / decision theory.
\item
  Decision-theoretic tools such as optimal control calculations rely on
  robust \emph{forecasting} more strongly than they rely on accurate
  \emph{mechanistic} relationships.
\item
  Adapting a non-parametric approach requires modification of existing
  methods for decision theory. We have illustrated how this might be
  done in the context of stochastic dynamic programming, opening the
  door for substantial further research into how these applications
  might be improved.
\item
  Anticipate improved relative performance in higher dimensional
  examples
\item
  Discuss constant escapement in model, in policies.
\item
  Limitations of this comparison: Are the maximum-likelihood solutions a
  straw man?
\item
  Discussion of alternative related approaches: POMDP/MOMDP,
\end{itemize}

@@

@@

@@

Geritz, Stefan a. H., and Eva Kisdi. 2011. ``Mathematical ecology: why
mechanistic models?'' \emph{Journal of mathematical biology} (dec).
doi:10.1007/s00285-011-0496-3.

@@

Polasky, Stephen, Stephen R. Carpenter, Carl Folke, and Bonnie Keeler.
2011. ``Decision-making under great uncertainty: environmental
management in an era of global change.'' \emph{Trends in ecology \&
evolution} (may): 1--7. doi:10.1016/j.tree.2011.04.007.

>>>>

Process-based models built on modern machine learning approaches offer a
powerful and underappreciated approach for ecological forecasting and
management. We present a novel example of how one such example -
non-parametric Gaussian processes, can be adapted to the context of
management and decision-making.

While effective machine learning algorithms have speckled the
literature, most theoretical ecologists in particular have been wary of
these `black-box' approaches in favor of simple, mechanistic models.
While a blind application of machine learning approaches would indeed be
dangerous grounds to stake either ecological understanding or management
practice, such concerns are not grounds to reject such methods out of
hand. Rather, further attention is warranted to identify the
circumstances in which machine-learning approaches will most likely
outperform existing more mechanistic approaches, or help crack difficult
management questions where ecological or dynamical complexity has caused
traditional approaches to founder.

Here we illustrate how a machine learning model can be implemented in
place of simple mechanistic model in a decision theory (optimal control)
problem, and highlight the features of this system that make the
nonparametric approach particularly effective.

Mechanistic models have long been the gold standard of theoretical
modeling in ecology (e.g.~see @Geritz2012 or @Cuddington2013). Only by
understanding the processes involved can we make reliable long term
predictions and build an knowledge of cause and effect that guides the
hypotheses we make, the data we collect, and the management decisions we
make. Process-based models, whether expressed in the language of
mathematics or English, identify the connection between mosquitoes and
the spread of malaria, or greenhouses gases and climate change, guiding
our approach to understand and manage these threats. Despite this
central importance, we argue here that ecologists would do well to give
greater attention to the role non-mechanistic models can play in
ecological management and decision making. The value of these approaches
is greatest in a context where decisions are made over short time
horizons, and updated as new data becomes available.

\subsection{The alternative to process-based models: pattern-based
modeling}

Before we proceed further it would be useful to define our terms and lay
some greater context for the discussion. Throughout, we will distinguish
``mechanistic'' or ``process-based'' models from models that are merely
``correlative'' or ``pattern-based''. Historically pattern-based
modeling meant regression (usually linear regression) and resided in
methodology rooted in statistics departments and was the primary focus
of empirical ecologists, while mechanistic modeling meant dynamical
systems (ODEs, PDEs, SDEs and their discrete kin), residing in
methodology from the mathematics department and was the primary focus of
theoretical ecologists.

We believe these historical divides continue to color much of the
literature today, with theorists more skeptical than empiricists of
statistical methodology and vice versa. Meanwhile, the ground underneath
has shifted. Originally pattern-matching approaches could be critiqued
on the grounds of their simplicity (not everything is linear) while
mechanistic approaches could be critiqued on their tenuous connections
to data -- model parameter values such as death rates, birth rates, etc.
would be estimated in advance and then stuck into the model, rather than
estimated in the context of the model itself. Today both of these
critiques are outdated. Computational power and hierarchical statistical
methods (particularly approximate likelihood or simulation techniques
such as particle filters and approximate Bayesian computing) have
brought the inference richer dynamical systems into their fold, while
pattern-based modeling has spawned an entirely new approach under the
banner of machine learning that has divided the statistics community
(See @Bremman2001, ``The Two Cultures''). Machine learning models can
represent almost arbitrarily complex patterns and incorporate learning
and decision making strategies that have earned them the name
`artificial intelligence.' Yet at the same time they can be far less
transparent their linear regression predecessors, making both astounding
successes and startling biases and failures like their biological
namesake. More ecologists, particularly theorists and modelers, would do
well to learn both how to take advantage of these strengths and
recognize their weaknesses.

Mechanistic modeling emphasizes the importance of capturing the correct
gross properties of a system over tracking minute fluctuations. For
instance, in selecting and parameterizing model of a population of
conservation concern, we may be most interested in getting the long-term
behavior correct -- such as identifying if the dynamics support
persistence of the population -- rather than worrying how well they
reflect the year-to-year fluctuations. We would certainly have good
reason to prefer such a model over alternatives which are irreconcilable
to the most basic biological processes, such as unbounded growth, or
growth curves that do not pass through the origin in a closed system. So
it may come as a surprise to realize that such obviously wrong models
can perform as well or even better than reasonable mechanistic models in
guiding ecological management and decision making.

Most management recommendations from the ecological literature are based
on (or at least motivated by) parametric models. Though in principle
these models reflect mechanistic underpinnings of biological
interactions involved, in practice real mechanisms are often unknown and
the true dynamics too complex to be captured by simple models
{[}@refs{]}. While simple mechanistic models can nevertheless provide
imporant insights into possible dynamics -- for instance, demonstrating
that a vaccine does not have to be 100\% effective to eliminate the
transmission of a virus {[}@Kermack1921{]} -- such approaches are not
well-suited for use in forecasting future dynamics upon which outcomes
management policy can be based. Despite a long history of work
emphasizing the difference between modeling for understanding (generic
or strategic models) and modeling for prediction (precise, specific, or
tactical models) (e.g. Levins 1966), simple parametric models continue
to find application in predictive management contexts. The management of
marine fisheries is a prime example, in which simple, mechanistically
motivated models of stock-recruitment relationships such as Ricker or
Beverton-Holt curves are systematically fit to data and used in solving
decision-theoretic or optimal control problems determining resource
management policies such as maximum sustainable yield {[}@refs{]}.

Stochastic dynamic programming @Mangel1982

To address structural uncertainty in the processes involved, such
mechanistic models can be adapted to accomidate uncertainty in the
parameters or consider a set of alternative models simultaneously
(belief SDP)

Further unknowns such as measurement uncertainty, parameter uncertainty,
unobserved states, knowledge of boundary conditions, etc. further
compound the issue. Though a hierarchical Bayesian approach provides a
natural way to address these from a statistical standpoint, formulating
reasonable parametric descriptions of each form of uncertainty is a
challenging task in itself, let alone the computational difficulties of
solving such a system. Cressie et al. (2009) provides a good account of
the successess and challenges of the approach. Applying these approaches
in the management context of sequential decision making, in which
forecasts must be obtained over a range of possible actions and updated
regularly as new information arrives makes such an approach less
feasible still.

Though machine learning approaches have begun to appear in the
ecological and conservation literature (Species distribution models),
including the Gaussian process based approach used here (Munch et al.
2005), they remain unfamiliar and untrusted approaches for most
ecologists. Machine learning approaches represent an essentially
pattern-based rather than process-based approach, raising the same
skepticism from most theoretical ecologists that they hold for older
correlative methods such as linear regression, while the complexity of
the techniques has barred their adoption in more empirical audiences.

A more immediate barrier to their adoption is the absence of a framework
for applying machine learning approaches to resource management
problems. Traditional approaches to optimal control (Pontryagin's
principle, stochastic dynamic programming) rely on knowledge of the
state equation, usually described by a simple parametric model. Here we
illustrate how a stochastic dynamic programming algorithm can
alternatively be driven by the predictions from a Gaussian process -- a
machine learning approximation to the state dynamics.\\

Management goals / decision-theoretic approaches need accurate
prediction over relevant (short?) timescales more than reasonable
long-term mechanisms. Machine-learning approaches may offer the benefit
of the hierarchical Bayesian approach without the practical and
computational limitations of their parametric kin. Non-parametric models
are flexible enough to take maximum advantage of the data available,
while being appropriately ambiguous about the dynamics of a system in
regions of parameter space that have been poorly or never sampled.

@@

The use of Gaussian process regression (known as Kreging in the
geospatial literature) to formulate a predictive model is relatively new
in the context of modeling dynamical systems {[}@Kocijan2005{]} and
introduced in the ecological modeling and fisheries management by Munch
et al. (2005). An accessible and thorough introduction to the
formulation and use of Gaussian processes can be found in
@Rasmussen2006.

The essense of the Gaussian process approach can be captured in the
following thought experiment: An exhaustive parametric approach to the
challenge of structural uncertainty might proceed by writing down all
possible functional forms for the underlying dynamical system with all
possible parameter values for each form, and then consider searching
over this huge space to select the most likely model and parameters; or
using a Bayesian approach, assign priors to each of these possible
models and infer the posterior distribution of possible models. The
Gaussian process approach can be thought of as a computationally
efficient approximation to this approach. Gaussian processes represent a
large class of models that can be though of as capturing or reasonably
approximating the set of models in this collection. By modeling at the
level of the process, rather than the level of parametric equation, we
can more concisely capture the possible behavior of these curves. In
place of a parametric model of the dynamical system, the Gaussian
Process approach postulates a prior distribution of (n-dimensional)
curves that can be though of as approximations to a range of possible
(parametric) models that might describe the data. The GP allows us to
consider a set of possible curves simultaneously.\\

\subsubsection{The optimal control problem}

@@

Non-parametric methods have received far too little attention in
ecological modeling efforts that are aimed at improved conservation
planning and decision making support.

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Adapting a non-parametric approach requires modification of existing
  methods for decision theory. We have illustrated how this might be
  done in the context of stochastic dynamic programming, opening the
  door for substantial further research into how these applications
  might be improved.
\end{itemize}

Three elements contribute to the machine learning model performing as
well or better than the mechanisitic models in this context.

\subsubsection{1. Relevant state space}

The dynamics occur over a range of state-space in which the machine
learning model closely matches the predictions of y implausible model
performs as well or better than the more plausible model.

This is the most obvious and immediate reason why the shortcomings that
appear to make the model implausible do not make it useless. This effect
is enhanced in our example because the management actions help drive the
system towards rather than away from this region of state-space.

\subsubsection{2. Predictive accuracy}

This application relies only on the predictive accuracy of the model,
not an interpretation of the parameters.

Predictive accuracy is not the goal of all modeling, as ecologists have
been observing for as long as they made models (perhaps none more
memorably than @Levins1969). Mechanistic modeling is at its most
powerful not when it is used to directly forecast future states but when
it provides an understanding of how to approach a problem. SIR-type
models from the epidemiological literature are a good example. While the
simplest SIR models have little predictive power over the outbreak
intensity or timing at a particular location, they provide a powerful
understanding of the spread of an infection in terms of a single,
biologically meaningful parameter: $R_0$, the basic reproductive number.
From the model, it becomes clear that management need not vaccinate
every member of the population to stop the spread, but rather it
suffices to vaccinate a sufficient fraction of the population to reduce
$R_0$ below 1.

\subsubsection{3. Time scale for new data}

In the sequential decision making problem we considered, we are
presented with new data after each action. The relevant timescale for
the prediction is thus not the long-term dynamics, which would be wildly
divergent, but the dynamics over this much shorter interval. While
ecologists may be hesitant to base continual management on a model with
obviously inaccurate long-term behavior, engineers tend to consider the
problem in frequency space and gravitate to the opposite position -- a
good control model should prioritize high-frequency accuracy over low
frequency accuracy. The differences in intuition may arise from the
timescales at which each profession can typically adjust their control
variables -- much faster for a control system of a chemical plant than
state policy for a natural resource. Still, the lesson is clear: when
facing repeated management decisions over a short timescale, such as
setting annual harvests of a fishery, it may be more valuable to use a
machine learning algorithm that makes accurate year-ahead predictions
that capture some of the high-frequency fluctuations that appear only as
noise in a mechanistic model of the long-term population dynamics.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Better express ambiguity in regions where data is not available
\end{enumerate}

One of the greatest strengths of mechanistic models is their greatest
weakness as well.

@@

While we have highlighted certain generic of this problem that allow the
nonparametric approach to excel -- short timescales between successive
observations and actions, the accuracy in the appropriate region of
state space, the ability to express uncertainty outside the observed
data -- there are equally several aspects for which the nonparametric
approach is at a relative disadvantage.

In this simulated example, the underlying dynamics are truly governed by
a simple parametric model, allowing the parametric approaches to be more
accurate. Similarly, because the dynamics are one-dimensional dynamics
and lead to stable nodes (rather than other attractors such as
limit-cycles resulting in oscillations), the training data provides
relatively limited information about the dynamics. For these reasons, we
anticipate that in higher-dimensional examples characteristic of
ecosystem management problems that the machine learning approach will
prove even more valuable.

The machine learning approach is also better suited to complex and
disparate data. Incorporating various sources of information into
mechanistic models can be an immensely difficult due to the increased
complexity involved. Only thanks to tandem advances in increasing
computational power and hierarchical statistical methodology have we
been able to tackle such intuitively important complexity (and the
potentially new available data that accompanies it) such as spatial
distribution, heterogeneities of space, time, and individuals, to shift
to ecosystem-based approaches from single-species based approaches.
Without the need to formulate mechanisms, many modern machine learning
algorithms can leverage potential information from all available sources
of data directly. The algorithms can recognize unanticipated or subtle
patterns in large data sets that enable more accurate predictions than
mechanistic models that are formulated at a more macroscopic level.

\begin{center}\rule{3in}{0.4pt}\end{center}

@@

\subsubsection{The Guassian process}

The Gaussian process is defined by a covariance kernel. By requiring our
kernel to follow a generic functional form, we can compactly describe
the Gaussian process using only a handful of parameters (Table 1)

\subsubsection{Formulating a dynamic programming solution}

@@

@@
